The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-08-26 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
More plots etc

While some success has been seen in filtering interference from observations \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, in which the data samples are cleaned from the contribution of interference, such an approach is often not feasible. Therefore, the most common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}. The consequence of the latter approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed. It is important to analyse the results of interference detection, as improving the understanding of RFI contamination can aid in several ways: understanding the effects on the science; improved observation scheduling; designing hardware and location of future telescopes with a maximal cost/benefit approach; and enforcing spectrum management strategies that are most beneficial for the observatories.

In this article, we will look specifically at the RFI situation of the MWA \citep{mwa}. The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31~MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison region in Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be build near the location of the MWA.

The interferometric arrays LOFAR, PAPER and the Long Wavelength Array (LWA) are all designed to perform 21-cm EoR science, and all observe in approximately the same frequency range as the MWA. Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI mitigation effectivity. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose any issues on regular observations despite that, unlike the MWA, LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of Epoch of Reionization long-integration observations with LOFAR. They reach near-thermal noise sensitivity, and argue that RFI is not a limitation for further increasing the sensitivity. Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}. This analysis concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down in longer integrations like Gaussian noise.

@@

\section{Method}
RFI detection (``data flagging'') is one of the first steps in the data processing. Several observatories or projects have designed pipelines that include RFI mitigation. In some projects, simple thresholding is used to mitigate the worst interference. For example, in \citet{parsons-paper-eorlimit-2014} a $3\sigma$ threshold is used for reducing PAPER data. However, more advanced techniques such as the \textsc{sumthreshold} technique have been shown to be more accurate \citep{post-correlation-rfi-classification}. A few pipelines that include RFI detection are \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR but used for various telescopes; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).

@@

To detect RFI in MWA observations, the \textsc{aoflagger} software is used. The \textsc{aoflagger} allows customizing the flagging strategies, such as the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of a different band-pass shape, different time and frequency resolutions and varying field of views. Specific strategies for several telescopes are implemented, including an MWA specific strategy which is used in this work. All \textsc{aoflagger} strategies are variations on the original LOFAR strategy. In this strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are flagged with the \textsc{sumthreshold} method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

@@

The \textsc{aoflagger} software provides a C++ library\footnote{\url{http://aoflagger.sourceforge.net/doc/api/}} that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It will also collect some statistics and write these into a measurement set with the same format as defined for LOFAR\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot these over various dimensions in an interactive manner.

Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data. MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. The first step for the scientist is to download the raw files for a given observation ID. There are various ways of processing MWA data afterwards. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not correct or convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format, and is subsequently imported into the astronomical software.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The \textsc{sumthreshold} technique and the scale-invariant rank operator that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency ranges that are handed over in one part to the flagger are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. This has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Therefore, we are working on a platform with sufficient memory that pre-runs \textsc{cotter} on all observations.

\begin{itemize}
 \item Do we mention miriad and CASA's flagging tools?
\end{itemize}

@@

\noindent\begin{figure*}
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
\caption{A very weakly observed transmitter centred on 156.66 MHz.}
\label{fig:156_7}
\end{center}
\end{figure*}

>>>>

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}. The consequence of the latter approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed. It is important to analyse the results of interference detection, as improving the understanding of RFI contamination can aid in several ways: understanding the effects on the science; improved observation scheduling; designing hardware and location of future telescopes with a maximal cost/benefit approach; and enforcing spectrum management strategies that are most beneficial for the observatories.

In this article, we will look specifically at the RFI situation of the MWA \citep{mwa}. The MWA is a low-frequency array consisting of 128 tiles, each with 16 dual-polarized dipoles, which allow observing between 80--300~MHz with a 31-MHz instantaneous bandwidth. Its main science driver is to detect redshifted 21-cm radio signals from the Epoch of Reionization (EoR; \citealt{bowman-science-with-the-mwa-2013}). To avoid as much RFI as possible, the MWA is located in the Murchison region in Western Australia. Analysing the interference situation of the MWA can improve the MWA observing and processing strategy, and will additionally also provide valuable information for the Square Kilometre Array (SKA), because the cores of the SKA low-frequency aperture arrays and SKA mid-frequency survey dishes are planned to be build near the location of the MWA.

The interferometric arrays LOFAR, PAPER and the Long Wavelength Array (LWA) are all designed to perform 21-cm EoR science, and all observe in approximately the same frequency range as the MWA. Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI mitigation effectivity. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose any issues on regular observations despite that, unlike the MWA, LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of Epoch of Reionization long-integration observations with LOFAR. They reach near-thermal noise sensitivity, and argue that RFI is not a limitation for further increasing the sensitivity. Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}. Those analyses concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down similar to Gaussian noise.

@@

\section{Method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. In some projects, simple thresholding is used to mitigate the worst interference. For example, in \citet{parsons-paper-eorlimit-2014} a $3\sigma$ threshold is used for reducing PAPER data. For more accurate detection, advanced techniques such as the \textsc{sumthreshold} technique have been developed \citep{post-correlation-rfi-classification}. Several observatories or projects have designed pipelines that include advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR but used for various telescopes; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}).

@@

To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying field of views. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are detected by the \textsc{sumthreshold} method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

@@

The \textsc{aoflagger} software provides a C++ library\footnote{\url{http://aoflagger.sourceforge.net/doc/api/}} that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. Additionally, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects some statistics and write these into a measurement set with the same format as used by LOFAR\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot these over various dimensions in an interactive manner.

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see TODO fix reference Ord et al. (2014). Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data. The first step for the scientist is to download the raw files for a given observation ID. There are various ways of processing MWA data afterwards. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not correct or convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format. The output is directly readable by the astronomical software.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The \textsc{sumthreshold} technique and the scale-invariant rank operator that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency ranges that are handed over in one part to the flagger are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. This has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Therefore, we are working on a platform with sufficient memory that pre-runs \textsc{cotter} on all observations. The partitioning of data to overcome the time-ordering problem is also use by the \textsc{ndppp} software used by LOFAR (TODO reference?).

@@

\noindent\begin{figure*}
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
\caption{A weakly-observed transmitter centred on 156.66 MHz.}
\label{fig:156_7}
\end{center}
\end{figure*}

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-stddev-per-set}\vspace{-1cm}
\caption{..}
\label{fig:stddev-per-set}
\end{center}
\end{figure*}

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set}\vspace{-1cm}
\caption{..}
\label{fig:rfi-per-set}
\end{center}
\end{figure*}
